Wait-free Queues with Polylogarithmic Step Complexity

### 2022-09-28 thesis/thesis.tex
review linearizability proof

\usepackage[letterpaper]{geometry}
\usepackage{tikz-qtree}
\usepackage{algorithm}
\usepackage{algpseudocode}
\makeatletter
\renewcommand{\ALG@beginalgorithmic}{\footnotesize}
\makeatother
\usepackage{graphicx}
\usepackage{subcaption}
%\usepackage{showkeys}
\usepackage{amsmath}

@@

\algdef{S}[FUNCTION]{Function}
   [3]{{\tt{\sl{#1}}} \textproc{\tt{#2}}\ifthenelse{\equal{#3}{}}{}{\tt{(#3)}}}

\algdef{E}[FUNCTION]{EndFunction}
   [1]{\algorithmicend\ \tt{\textproc{#1}}}

\algrenewcommand\Call[2]{\tt{\textproc{#1}\ifthenelse{\equal{#2}{}}{}{(#2)}}}

@@

A process may wish to know the $i$th propagated operation or the rank of a propagated operation in the linearization. In our case of implementing a queue, we can make an assumption that one process only wishes to know the rank of a dequeue and one tries to get an enqueue with its rank. \nf{enqueue}s and \nf{dequeue}s are appended to the tree and when we want to find the response to a \nf{dequeue}, we compute the place of the dequeue in the linearization and using the rank of the dequeue among deqeueues and some information stored in the root we compute which enqueue is the answer to the dequeue or if the answer is null. If the answer was some enqueue we find the enqueue using \texttt{DSearch(i)} and \nf{GetENQ(n,b,i)}. \texttt{DSearch(i)} finds the block containing the $i$th enqueue in the root and \nf{GetENQ(n,b,i)} finds its sub-block recursively to reach a leaf. \texttt{Index()} is similar but more complicated, finding super-blocks from a leaf to the root. The main challenge in each level of \texttt{Get(i)} and \texttt{Index(op)} is that it should take polylogarithmic steps with respect to $p$. After appending operation \texttt{op} to the root, processes can find out information about the linearization ordering using \texttt{Get(i)} and \texttt{Index(op)}. Each block stores an extra constant amount of information (like prefix sums) to allow binary searches to find the required block in a node quickly.

@@

\Function{Object}{Dequeue()}{} \cmt{Creates a block with null value element, appends it to the tree, computes its order among operations, and returns its response.}
\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}} 
\State \tt{newBlock.element= null}
\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.\head].sum\sub{enq}}
\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.\head].sum\sub{deq}+1}
\State \tt{leaf.}\Call{Append}{newBlock}
\State \tt{<b, i>=} \Call{IndexDeq}{leaf.\head, 1}
\State \tt{output=} \Call{FindResponse}{b, i} 
\label{deqRest}
\State \Return{\tt{output}}
\EndFunction{Dequeue}

@@

\Function{<int, int>}{FindResponse}{\sl{int} b, \sl{int} i}\cmt{Returns the the response to the $D_{root,b,i}$.}
\If{\tt{ root.blocks[b-1].\size}\tt{ + root.blocks[b].num\sub{enq} - i $<$ 0}} \cmt{Check if the queue is empty.}
\State \Return \tt{null} \label{checkEmpty}
\Else
\State \tt{e= i - root.blocks[b-1].size + root.blocks[b-1].sum\sub{enq}} \label{computeE}
\cmt{$E_e(root)$ is the response.}
\State \Return \tt{root.GetENQ(root.\Call{DSearch}{e, b})}\label{findAnswer}
\EndIf
\EndFunction{FindResponse}

@@

\Statex $\leadsto$ \textsf{Precondition:~\tt{blocks[b].num\sub{enq}$\geq$i$\geq 1$}}
\Function{element}{GetEnq}{\sl{int} b, \sl{int} i} \cmt{Returns the element of $E_i(this,b)$.}
\If{\tt{this} \bf{is} \tt{leaf}}
\State\Return \tt{blocks[b].element} \label{getBaseCase}
\ElsIf{\tt{i $\leq$ blocks[b].num\sub{enq-left}}} \label{leftOrRight} \cmt{$E_i(this,b)$ is in the left child of this node.}
\State \tt{subBlock= left.\Call{BinarySearch}{sum\sub{enq}, i+blocks[b-1].sum\sub{enq-left}, blocks[b-1].end\sub{left}+1, blocks[b].end\sub{left}}} \label{leftChildGet}
\State \Return\tt{left.}\Call{GetEnq}{subBlock, i} 
\Else
\State \tt{i= i-blocks[b].num\sub{enq-left}}
\State\tt{subBlock= right.\Call{BinarySearch}{sum\sub{enq}, i+right.blocks[b-1].sum\sub{enq-right}, blocks[b-1].end\sub{right}+1, blocks[b].end\sub{right}}} \label{rightChildGet}
\State \Return\tt{right.}\Call{GetEnq}{subBlock, i} 
\EndIf
\EndFunction{GetEnq}

\Statex
\Statex $\leadsto$ \textsf{Precondition: \tt{b}th block of the node has propagated up to the root and \tt{blocks[b].num\sub{enq}$\geq$i}.}
\Function{<int, int>}{IndexDeq}{\sl{int} b, \sl{int} i} \cmt{Returns \tt{<x, y>} if $D_{this,b,i}=D_{root,x,y}$.}
\If{\tt{this} \bf{is} \tt{root}}
\State\Return \tt{<b, i>} \label{indexBaseCase}
\Else
\State \tt{dir= (parent.left==n)? left: right} \cmt{check if this node is a left or a right child}
\State \tt{superBlock= parent.\Call{BinarySearch}{sum\sub{deq-dir}, i+blocks[b-1].sum\sub{deq}, blocks[b].super, blocks[b].super+1}} \label{computeSuper}

@@

\If{\tt{dir {\keywordfont is} right}} \label{considerRight}
\State \tt{i+= blocks[b-1].sum\sub{enq}-blocks[superBlock-1].sum\sub{enq-right}} \cmt{consider the enqueues in the previous blocks from the right child} \label{considerPreviousRight}
\State \tt{i+= blocks[superBlock].num\sub{deq-left}} \cmt{consider the dequeues from the right child} \label{considerLeftBeforeRight}
\EndIf \label{computeISuperEnd}
\State \Return\tt{this.parent.}\Call{IndexDeq}{superBlock, i}
\EndIf
\EndFunction{IndexDeq}

@@

We adopt linearizability as our definition of correctness. In our case, where we create the linearization ordering in the root, we need to prove (1) the ordering is legal, i.e, for every execution on our queue if operation $op_1$ terminates before operation $op_2$ then $op_1$ is linearized before operation $op_2$ and (2) if we do operations sequentially in their the linearization order, operations get the same results as in our queue. The proof is structured like this. First, we define and prove some facts about blocks and the node's \nf{head} field. Then, we introduce  the linearization ordering formally. Next, we prove double \nf{Refresh} on a node is enough to propagate its children's new operations up to the node, which is used to prove (1). After this, we prove some claims about the size and operations of each block, which we use to prove the correctness of \nf{DSearch()}, \nf{GetEnq()} and \nf{IndexDeq()}. Finally, we prove the correctness of the way we compute the response of a dequeue, which establishes (2).

@@

\begin{invariant}[headPosition] \label{lem::headPosition} If the value of $n.\nf{head}$ is $h$ then $\nf{$n$.blocks[$i$]}=\nf{null}$ for ${i>h}$ and $\nf{$n$.blocks[$i$]}\neq\nf{null}$ for ${0\leq i<h}$.
\end{invariant}
\begin{proof}
Initially the invariant is true since $n.\nf{head}=1$, $n.\nf{blocks[0]}\neq\nf{null}$  and $n.\nf{blocks[$x$]}=\nf{null}$ for every $x>0$. The truth of the invariant may be affected by writing into \nf{n.blocks} or incrementing \nf{n.head}. We show the invariant still holds after such changes.

In the algorithm, $n.\nf{blocks}$ is modified only on Line \ref{cas}, which updates $n.\nf{blocks[$h$]}$ where $h$ is the value read from $n.\nf{head}$ in Line \ref{readHead}. Since the \nf{CAS} in Line \ref{cas} is successful it means $n.\nf{head}$ has not changed from $h$ before doing the \nf{CAS}: if $n$\nf{.head} had changed before the \nf{CAS} then $n$\nf{.blocks[}$h$\nf{]}$\neq$\nf{null} and by the induction hypothesis, so the \nf{CAS} would fail. Writing into \nf{$n$.blocks[$n$.head$=h$]} preserves the invariant, since the claim does not talk about the content of \nf{$n$.blocks[$n$.head]}.

@@

\pagebreak
We define the subblocks of a block recursively.

@@

@@

@@

We show $EST_{n\nf{.left}}^{t^R}= ops(n\nf{.left.blocks[0..n.left.head$_{309}-1$]}) \subseteq ops(n.blocks_{\ref{cas}}) = ops(n\nf{.blocks[0..}n.head_{\ref{cas}}\nf{]})$.
%As \nf{CAS} in Line \ref{cas} returns \nf{true} the block \nf{new} is written into \nf{n.blocks[$n.head_{\ref{readHead}}$]}. By the Definition \ref{def::subblock} for the Subblock  the block \nf{new} contains \nf{n.left.blocks[n.blocks[$n.head_{\ref{readHead}}-1$].end\sub{left}+1..$n.left.head_{\ref{lastLine}}-1$]} (see Lines \ref{prevLine} and \ref{lastLine}).
Line \ref{cas} stores a block \nf{new} in $n$ that has $\nf{end\sub{left}}=n.left.head_{\ref{lastLine}}-1$. Therefore by Definition \ref{def::subblock}, after the successful \nf{CAS} in Line \ref{cas} we know all blocks in \nf{n.left.blocks[1..$n.left.head_{\ref{lastLine}}-1$]} are subblocks of \nf{n.blocks[1..$n.head_{\ref{readHead}}$]}. Because of Lemma \ref{nonDecreasingHead} we have $n.left.head_{309}-1<n.left.head_{\ref{lastLine}}-1$ and $n.head_{\ref{readHead}}<n.head_{\ref{cas}}$. From Observation \ref{lem::blocksOrder} the claim follows.  The proof for the right child is the same.
\end{proof}
\begin{figure}[hbt]
  \center\includegraphics[width=5in]{pics/trueRefresh}
  \caption{New established operations of the left child are in the new block. (TO UPDATE)}
\end{figure}

@@

\begin{corollary}
For $A=\nf{Append(}b\nf{)}$ we have ${ops(b)} \subseteq ops(EST_{\nf{n}, t^A})$ where $n\in\{\text{nodes in the path from the leaf to the root}\}$.
\end{corollary}
\begin{proof}
$A$ adds $b$ to the assigned leaf of the process, establishes it at Line \ref{appendEnd} and then calls \nf{Propagate()} on the parent of the leaf where it appended $b$. For every node $n$, $n.$\nf{Propagate()} appends $b$ to $n$, establishes it by Corollary \ref{lem::propagateStep} and then calls $n.$\nf{parent.Propagate()} utill $n$ is \nf{root}.
\end{proof}

@@

Now we prove some claims about the size and operations of a block. These lemmas will be used for analysis and correctness of \nf{GetEnq()} and \nf{IndexDeq()}.

@@

\begin{lemma}[Get correctness] \label{get}
If $\nf{$n$.blocks[$b$].num\sub{enq}} \geq i$ then \nf{$n$.GetENQ($b$, $i$)} returns $E_i(n,b)$\nf{.element}.
\end{lemma}
\begin{proof}
We are going to prove this lemma by induction on the height of node \nf{n}. For the base case, $n$ is a leaf. Leaf blocks each contain exactly one operation, so only \nf{n.GetENQ(b,1)} can be called when $n$ is a leaf. Line \ref{getBaseCase} of \nf{$n$.GetENQ($b$, $1$)} returns the \nf{element} of the \nf{enqueue} operation stored in the $b$th block of leaf \nf{n} .

For the induction step we prove if \nf{$n$.child.GetENQ($sb$, i)} returns $E_i(n.child, sb)$ then \nf{$n$.GetENQ($b$, $i$)} returns $E_i(n,b)$.  From Definition  \ref{ordering} of $E(n,b)$ operations from the left subblocks come before the operations from the right subblocks in a block (See Figure \ref{figGet}). \nf{num\sub{enq-left}} field  in \nf{$n$.blocks[$b$]} stores the number of \nf{Enqueue()} operations from the blocks's subblocks in the left child of \nf{n}. So the $i$th enqueue operation is propagated from the right child if and only if $i$ is greater than \nf{$b$.num\sub{enq-left}}. Line \ref{leftOrRight} decides whether the \nf{i}th enqueue in  \nf{b}th block  of internal node \nf{n} is in the  left child or right child subblocks of \nf{n.blocks[b]}. By definition \ref{def::ops} to find an operation in subblocks of \nf{$n$.blocks[$i$]} we need to search in the range \texttt{$n$.left.blocks[$n$.blocks[$i$-1].end\textsubscript{left}+1..$n$.blocks[$i$].end\textsubscript{left}]} $\cup$ \texttt{$n$.right.blocks[$n$.blocks[$i$-1].end\textsubscript{right}+1..$n$.blocks[$i$].end\textsubscript{right}]}.

 There are $eb=n.blocks[b-1].sum_{enq-left}$ enqueues in the blocks before the left subblocks of $n$\nf{.blocks[$b$]}, so $E_i(n,b)$ is $E_{i+eb}(n.left)$ which is $E_{i^\prime}(n.left,b^\prime)$ for some $b^\prime$ and $i^\prime$. We can compute $b^\prime$ and then search for $i+eb$th enqueue in \tt{n.left}, where $i^\prime$ is \tt{i+eb-n.left.blocks[}$b^\prime-1$\tt{].sum\sub{enq}}. The parameters in Line \ref{leftChildGet} are for searching $E_{i+eb}(n.left)$ in \nf{n.left.block} in the expected range of blocks, so this \nf{BinarySearch} returns the index of the subblock containing $E_i(n,b)$.

Otherwise the enqueue we are looking for is in the right child. Because left \nf{Enqueue}s are before the right \nf{Enqueues}, there are \nf{n.blocks[b].num\sub{enq-left}} enqueues ahead from the left child. So we need to search for \nf{i-n.blocks[b].num\sub{enq-left}+ n.blocks[b-1].sum\sub{enq-right}} (Line \ref{rightChildGet}). Other parameters for the left child are chosen similar to the right child.

So, in both cases the direct subblock containing $E_i(n,b)$ is computed in Lines \ref{leftChildGet} and \ref{rightChildGet}.
 Finally, \nf{n.child.GetENQ(subblock, i)} is invoked on the subblock containing $E_i(n,b)$ and it returns $E_i(n,b)$ by the hypothesis of the induction.
\end{proof}

@@

%\pagebreak
%\begin{lemma}[Computing SuperBlock]\label{superBlock}
%For the \nf{superblock} value computed in line \ref{computeSuper} of \nf{n.IndexDEQ(b,i)} we have \nf{n.parent.blocks[superblock]} contains $D_{n,b,i}$.
%\end{lemma}
%\begin{proof}
%First we show the value read for \nf{super[b.group]} in line 418 is not null.
%  Values \nf{np\textsubscript{dir}} read in lines \ref{setNP}, \nf{super} are set before incrementing in lines \ref{setSuper},\ref{incNP}. So before incrementing \nf{num\sub{propagated}, super[num\sub{propagated}]} is set so it cannot be null while reading. Then by Lemma \ref{superRange}if we search in the range $p$, we can find the superblock.
%
%\end{proof}
Now using Corollary \ref{superRelation} on each step of the \nf{IndexDEQ} we prove its correctness.

\begin{lemma}[\nf{IndexDEQ} correctness]
 If $1\leq i \leq\nf{$n$.blocks[$b$].num\sub{deq}}$ then \nf{$n$.IndexDEQ($b$,$i$)} returns the rank in $D(root)$ of $D_i(n,b)$.
\end{lemma}
\begin{proof}
We will prove this by induction on the distance of $n$ from the \nf{root}. The base case where $n$ is \nf{root} is trivial (Line \ref{indexBaseCase}).
For the non-root nodes \nf{$n$.IndexDEQ($b$, $i$)} computes the superblock of the $b$th block of $n$ in Line \ref{computeSuper} by Corollary \ref{superRelation}. So preconditions of the invocation of \nf{BinarySearch} is satisfied. After that the order of $D_i(n,b)$ in $D(n.parent, superblock)$ is computed in Lines \ref{computeISuperStart}--\ref{computeISuperEnd}. Note that by Lemma \ref{blockSize} in each block there is at most one \nf{Dequeue} from each process and by Definition \ref{ordering} \nf{Dequeue}s in a block are ordered based on the order of its subblocks  from left to right. Then if $D_i(n,b)$ was propagated from the right child, the number of dequeues in the subblocks from the left child is considered to be ahead of the computed index (Line \ref{considerRight}). Finally \nf{IndexDEQ} is called on \nf{$n$.parent} recursively and it returns the response from induction hypothesis. 
\end{proof}

\pagebreak
We have explained the picture of proof in the first paragraph of proof section. Should I repeat here?

\begin{lemma}[satisfiability] \label{linearSat}
$L$ can be a linearization ordering.
\end{lemma}
\begin{proof}
A linearization for an execution is legal if, $op_{1}$ terminates before $op_{2}$ starts in execution then $op_{1}$ is before $op_{2}$ in the linearization. If $op_{1}$ terminates before $op_{2}$ starts,$op_{1}$\nf{.Append()} has terminated before $op_{2}$\nf{.Append()} started. From Lemma \ref{append} $op_{1}$ is in \nf{root.blocks} before $op_{2}$ starts to propagate. By definition of $L$, $op_{1}$ is linearized before $op_{2}$.
\end{proof}
Once some operations are aggregated in one block they will get propagated up to the root together and they can linearized in any order among themselves. The way we have chosen to order the operations in a block makes \nf{GetENQ}, \nf{FindResponse} efficent.

\begin{definition}\label{defNullDeq}
If a \nf{Dequeue} operation returns null it is called \it{null dequeue}, otherwise it is called \it{non-null dequeue}.
\end{definition}

\begin{observation} \label{responseToADeq}
In a sequential execution on a queue, $k$th non-null dequeue returns the \nf{element} of $k$th enqueue.
\end{observation}

>>>>

\usepackage[letterpaper, margin=1in]{geometry}
\usepackage{tikz-qtree}
\usepackage{algorithm}
\usepackage{algpseudocode}
\makeatletter
\renewcommand{\ALG@beginalgorithmic}{\footnotesize}
\makeatother
\usepackage{graphicx}
\usepackage{subcaption}
%\usepackage{showkeys}
\usepackage{amsmath}

@@

\algdef{S}[FUNCTION]{Function}
   [3]{{\tt{\sl{#1}}} {\tt{#2}}\ifthenelse{\equal{#3}{}}{}{\tt{(#3)}}}

\algdef{E}[FUNCTION]{EndFunction}
   [1]{\algorithmicend\ \tt{{#1}}}

\algrenewcommand\Call[2]{\tt{{#1}\ifthenelse{\equal{#2}{}}{}{(#2)}}}

@@

A process may wish to know the $i$th propagated operation or the rank of a propagated operation in the linearization. In our case of implementing a queue, we can make an assumption that one process only wishes to know the rank of a dequeue and one tries to get an enqueue with its rank. \nf{enqueue}s and \nf{dequeue}s are appended to the tree and when we want to find the response to a \nf{dequeue}, we compute the place of the dequeue in the linearization and using the rank of the dequeue among deqeueues and some information stored in the root we compute which enqueue is the answer to the dequeue or if the answer is null. If the answer was some enqueue we find the enqueue using \texttt{DSearch(i)} and \nf{GetEnqueue(n,b,i)}. \texttt{DSearch(i)} finds the block containing the $i$th enqueue in the root and \nf{GetEnqueue(n,b,i)} finds its sub-block recursively to reach a leaf. \texttt{Index()} is similar but more complicated, finding super-blocks from a leaf to the root. The main challenge in each level of \texttt{Get(i)} and \texttt{Index(op)} is that it should take polylogarithmic steps with respect to $p$. After appending operation \texttt{op} to the root, processes can find out information about the linearization ordering using \texttt{Get(i)} and \texttt{Index(op)}. Each block stores an extra constant amount of information (like prefix sums) to allow binary searches to find the required block in a node quickly.

@@

\Function{Object}{Dequeue()}{} \cmt{Creates a block with null value element, appends it to the tree, computes its order among operations, and returns its response.}
\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}} 
\State \tt{newBlock.element= null}
\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.\head].sum\sub{enq}}
\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.\head].sum\sub{deq}+1}
\State \tt{leaf.}\Call{Append}{newBlock}
\State \tt{<b, i>=} \Call{IndexDequeue}{leaf.\head, 1}
\State \tt{output=} \Call{FindResponse}{b, i} 
\label{deqRest}
\State \Return{\tt{output}}
\EndFunction{Dequeue}

@@

\Function{<int, int>}{FindResponse}{\sl{int} b, \sl{int} i}\cmt{Returns the the response to the $D_{root,b,i}$.}
\If{\tt{ root.blocks[b-1].\size}\tt{ + root.blocks[b].num\sub{enq} - i $<$ 0}} \label{checkEmpty}\cmt{Check if the queue is empty.}
\State \Return \tt{null} \label{returnNull}
\Else
\State \tt{e= i - root.blocks[b-1].size + root.blocks[b-1].sum\sub{enq}} \label{computeE}
\cmt{$E_e(root)$ is the response.}
\State \Return \tt{root.GetEnqueue(root.\Call{DSearch}{e, b})}\label{findAnswer}
\EndIf
\EndFunction{FindResponse}

@@

\Statex $\leadsto$ \textsf{Precondition:~\tt{blocks[b].num\sub{enq}$\geq$i$\geq 1$}}
\Function{element}{GetEnqueue}{\sl{int} b, \sl{int} i} \cmt{Returns the element of $E_i(this,b)$.}
\If{\tt{this} \bf{is} \tt{leaf}}
\State\Return \tt{blocks[b].element} \label{getBaseCase}
\ElsIf{\tt{i $\leq$ blocks[b].num\sub{enq-left}}} \label{leftOrRight} \cmt{$E_i(this,b)$ is in the left child of this node.}
\State \tt{subBlock= left.\Call{BinarySearch}{sum\sub{enq}, i+blocks[b-1].sum\sub{enq-left}, blocks[b-1].end\sub{left}+1, blocks[b].end\sub{left}}} \label{leftChildGet}
\State \Return\tt{left.}\Call{GetEnqueue}{subBlock, i} 
\Else
\State \tt{i= i-blocks[b].num\sub{enq-left}}
\State\tt{subBlock= right.\Call{BinarySearch}{sum\sub{enq}, i+right.blocks[b-1].sum\sub{enq-right}, blocks[b-1].end\sub{right}+1, blocks[b].end\sub{right}}} \label{rightChildGet}
\State \Return\tt{right.}\Call{GetEnqueue}{subBlock, i} 
\EndIf
\EndFunction{GetEnqueue}

\Statex
\Statex $\leadsto$ \textsf{Precondition: \tt{b}th block of the node has propagated up to the root and \tt{blocks[b].num\sub{enq}$\geq$i}.}
\Function{<int, int>}{IndexDequeue}{\sl{int} b, \sl{int} i} \cmt{Returns \tt{<x, y>} if $D_{this,b,i}=D_{root,x,y}$.}
\If{\tt{this} \bf{is} \tt{root}}
\State\Return \tt{<b, i>} \label{indexBaseCase}
\Else
\State \tt{dir= (parent.left==n)? left: right} \cmt{check if this node is a left or a right child}
\State \tt{superBlock= parent.\Call{BinarySearch}{sum\sub{deq-dir}, i+blocks[b-1].sum\sub{deq}, blocks[b].super, blocks[b].super+1}} \label{computeSuper}

@@

\If{\tt{dir {\keywordfont is} right}} \label{considerRight}
\State \tt{i+= blocks[b-1].sum\sub{enq}-blocks[superBlock-1].sum\sub{enq-right}} \cmt{consider the enqueues in the previous blocks from the right child} \label{considerPreviousRight}
\State \tt{i+= blocks[superBlock].num\sub{deq-left}} \cmt{consider the dequeues from the right child} \label{considerLeftBeforeRight}
\EndIf \label{computeISuperEnd}
\State \Return\tt{this.parent.}\Call{IndexDequeue}{superBlock, i}
\EndIf
\EndFunction{IndexDequeue}

@@

We adopt linearizability as our definition of correctness. In our case, where we create the linearization ordering in the root, we need to prove (1) the ordering is legal, i.e, for every execution on our queue if operation $op_1$ terminates before operation $op_2$ then $op_1$ is linearized before operation $op_2$ and (2) if we do operations sequentially in their the linearization order, operations get the same results as in our queue. The proof is structured like this. First, we define and prove some facts about blocks and the node's \nf{head} field. Then, we introduce  the linearization ordering formally. Next, we prove double \nf{Refresh} on a node is enough to propagate its children's new operations up to the node, which is used to prove (1). After this, we prove some claims about the size and operations of each block, which we use to prove the correctness of \nf{DSearch()}, \nf{GetEnqueue()} and \nf{IndexDequeue()}. Finally, we prove the correctness of the way we compute the response of a dequeue, which establishes (2).

\subsection{Basic Properties}
In this subsection we talk about some properties of blocks and fields of the tree nodes.

@@

\begin{invariant}[headPosition] \label{lem::headPosition} If the value of $n.\nf{head}$ is $h$ then $\nf{$n$.blocks[$i$]}=\nf{null}$ for ${i>h}$ and $\nf{$n$.blocks[$i$]}\neq\nf{null}$ for ${0\leq i<h}$.
\end{invariant}
\begin{proof}
Initially the invariant is true since $n.\nf{head}=1$, $n.\nf{blocks[0]}\neq\nf{null}$  and $n.\nf{blocks[$x$]}=\nf{null}$ for every $x>0$. The truth of the invariant may be affected by writing into \nf{n.blocks} or incrementing \nf{n.head}. We show that if the invariant holds before such a change  then it  still holds after the change.

In the algorithm, $n.\nf{blocks}$ is modified only on Line \ref{cas}, which updates $n.\nf{blocks[$h$]}$ where $h$ is the value read from $n.\nf{head}$ in Line \ref{readHead}. Since the \nf{CAS} in Line \ref{cas} is successful it means $n.\nf{head}$ has not changed from $h$ before doing the \nf{CAS}: if $n$\nf{.head} had changed before the \nf{CAS} then it would be greater than $h$ by Observation \ref{nonDecreasingHead} and hence $n$\nf{.blocks[}$h$\nf{]}$\neq$\nf{null} and by the induction hypothesis, so the \nf{CAS} would fail. Writing into \nf{$n$.blocks[$h$]} when $h=n$\nf{.head} preserves the invariant, since the claim does not talk about the content of \nf{$n$.blocks[$n$.head]}.

@@

We define the subblocks of a block recursively.

@@

\subsection{Ordering Operations}

@@

\subsection{Propagating Operations to the Root}
In this section we explain why two \nf{Refresh}es are enough to propagate a nodes operations to its parent.

@@

We show $EST_{n\nf{.left}}^{t^R}= ops(n\nf{.left.blocks[0..n.left.head$_{309}-1$]}) \subseteq ops(n.blocks_{\ref{cas}}) = ops(n\nf{.blocks[0..}n.head_{\ref{cas}}\nf{]})$.

Line \ref{cas} stores a block \nf{new} in $n$ that has $\nf{end\sub{left}}=n.left.head_{\ref{lastLine}}-1$. Therefore by Definition \ref{def::subblock}, after the successful \nf{CAS} in Line \ref{cas} we know all blocks in \nf{n.left.blocks[1..$n.left.head_{\ref{lastLine}}-1$]} are subblocks of \nf{n.blocks[1..$n.head_{\ref{readHead}}$]}. Because of Lemma \ref{nonDecreasingHead} we have $n.left.head_{309}-1<n.left.head_{\ref{lastLine}}-1$ and $n.head_{\ref{readHead}}<n.head_{\ref{cas}}$. From Observation \ref{lem::blocksOrder} the claim follows.  The proof for the right child is the same.
\end{proof}
\begin{figure}[hbt]
  \center\includegraphics[width=5in]{pics/trueRefresh}
  \caption{New established operations of the left child are in the new block. (TO UPDATE)}
\end{figure}

@@

\begin{corollary}\label{lem::append}
For $A=\nf{Append(}b\nf{)}$ we have ${ops(b)} \subseteq ops(EST_{\nf{n}, t^A})$ where $n\in\{\text{nodes in the path from the leaf to the root}\}$.
\end{corollary}
\begin{proof}
$A$ adds $b$ to the assigned leaf of the process, establishes it at Line \ref{appendEnd} and then calls \nf{Propagate()} on the parent of the leaf where it appended $b$. For every node $n$, $n.$\nf{Propagate()} appends $b$ to $n$, establishes it by Corollary \ref{lem::propagateStep} and then calls $n.$\nf{parent.Propagate()} utill $n$ is \nf{root}.
\end{proof}

@@

\subsection{Correctness of GetEnqueue and IndexDequeue}

Now we prove some claims about the size and operations of a block. These lemmas will be used for analysis and correctness of \nf{GetEnqueue()} and \nf{IndexDequeue()}.

@@

\begin{lemma}[\nf{GetEnqueue} correctness] \label{get}
If $1\leq i \leq \nf{$n$.blocks[$b$].num\sub{enq}}$ then \nf{$n$.GetEnqueue($b$, $i$)} returns $E_i(n,b)$\nf{.element}.
\end{lemma}
\begin{proof}
We are going to prove this lemma by induction on the height of node \nf{n}. For the base case, suppose $n$ is a leaf. Leaf blocks each contain exactly one operation, so only \nf{$n$.GetEnqueue(b,1)} can be called when $n$ is a leaf. Line \ref{getBaseCase} of \nf{$n$.GetEnqueue($b$, $1$)} returns the \nf{element} of the \nf{Enqueue} operation stored in the $b$th block of leaf ${n}$.

For the induction step we prove if \nf{$n$.child.GetEnqueue($sb$, i)} returns $E_i(n.child, sb)$ then \nf{$n$.GetEnqueue($b$, $i$)} returns $E_i(n,b)$.  From Definition \ref{ordering} of $E(n,b)$, operations from the left subblocks come before the operations from the right subblocks in a block (see Figure \ref{figGet}). \nf{num\sub{enq-left}} field  in \nf{$n$.blocks[$b$]} is the number of \nf{Enqueue()} operations from the blocks's subblocks in the left child of \nf{n}. So the $i$th enqueue operation is propagated from the right child if and only if $i$ is greater than \nf{$b$.num\sub{enq-left}}. Line \ref{leftOrRight} decides whether the \nf{i}th enqueue in  \nf{b}th block  of internal node \nf{n} is in the  left child or right child subblocks of \nf{n.blocks[b]}. By Definitions \ref{def::ops}, \ref{def::subblock} to find an operation in subblocks of \nf{$n$.blocks[$i$]} we need to search in the range
\begin{align*}
&\texttt{$n$.left.blocks[$n$.blocks[$i$-1].end\textsubscript{left}+1..$n$.blocks[$i$].end\textsubscript{left}]}  \cup \\
&\texttt{$n$.right.blocks[$n$.blocks[$i$-1].end\textsubscript{right}+1..$n$.blocks[$i$].end\textsubscript{right}]}.
 \end{align*}
First we consider the case where the \nf{Enqueue} we are looking is in the left child. There are $eb=n.\nf{blocks[}b-1\nf{].sum\sub{enq-left}}$ enqueues in the blocks before the left subblocks of $n$\nf{.blocks[$b$]}, so $E_i(n,b)$ is $E_{i+eb}(n.left)$ which is $E_{i^\prime}(n.left,b^\prime)$ for some $b^\prime$ and $i^\prime$. We can compute $b^\prime$ and then search for the $i+eb$th enqueue in \tt{n.left}, where $i^\prime$ is \tt{i+eb-n.left.blocks[}$b^\prime-1$\tt{].sum\sub{enq}}. The parameters in Line \ref{leftChildGet} are for searching $E_{i+eb}(n.left)$ in \nf{n.left.block} in the range of left subblocks of $n$\nf{.blocks[}$b$\nf{]}, so this \nf{BinarySearch} returns the index of the subblock containing $E_i(n,b)$.

Otherwise, the enqueue we are looking for is in the right child. Because \nf{Enqueue}s from the left subblocks are ordered before the \nf{Enqueue}s from the right subblocks, there are \nf{n.blocks[b].num\sub{enq-left}} enqueues ahead of $E_i(n,b)$ from the left child. So we need to search for $i-n\nf{.blocks[}b\nf{].num\sub{enq-left}}+ n\nf{.blocks[}b-1\nf{].sum\sub{enq-right}}$ (Line \ref{rightChildGet}). Other parameters for the right child are chosen similarly to the left child.

So, in both cases the direct subblock containing $E_i(n,b)$ is computed in Line \ref{leftChildGet} or \ref{rightChildGet}.
 Finally, \nf{n.child.GetEnqueue(subblock, i)} is invoked on the subblock containing $E_i(n,b)$ and it returns $E_i(n,b)$ by the hypothesis of the induction.
\end{proof}

@@

Now using Corollary \ref{superRelation} on each step of the \nf{IndexDequeue} we prove its correctness.

\begin{lemma}[\nf{IndexDequeue} correctness]
 If $1\leq i \leq\nf{$n$.blocks[$b$].num\sub{deq}}$ then \nf{$n$.IndexDequeue($b$,$i$)} returns the rank in $D(root)$ of $D_i(n,b)$.
\end{lemma}
\begin{proof}
We will prove this by induction on the distance of $n$ from the \nf{root}. The base case where $n$ is \nf{root} is trivial (Line \ref{indexBaseCase}).
For the non-root nodes \nf{$n$.IndexDequeue($b$, $i$)} computes the superblock of the $b$th block of $n$ in Line \ref{computeSuper} by Corollary \ref{superRelation}. So preconditions of the invocation of \nf{BinarySearch} is satisfied. After that the order of $D_i(n,b)$ in $D(n.parent, superblock)$ is computed in Lines \ref{computeISuperStart}--\ref{computeISuperEnd}. Note that by Lemma \ref{blockSize} in each block there is at most one \nf{Dequeue} from each process and by Definition \ref{ordering} \nf{Dequeue}s in a block are ordered based on the order of its subblocks  from left to right. Then if $D_i(n,b)$ was propagated from the right child, the number of dequeues in the subblocks from the left child is considered to be ahead of the computed index (Line \ref{considerRight}). Finally \nf{IndexDequeue} is called on \nf{$n$.parent} recursively and it returns the response from induction hypothesis. 
\end{proof}

\pagebreak
\subsection{Linearizability}
We now prove the two properties needed for linearizability.

\begin{lemma} \label{linearSat}
$L$ is a legal linearization ordering.
\end{lemma}
\begin{proof}
A linearization for an execution is legal if, $op_{1}$ terminates before $op_{2}$ starts in execution then $op_{1}$ is before $op_{2}$ in the linearization. If $op_{1}$ terminates before $op_{2}$ starts,  $op_{1}$\nf{.Append()} has terminated before $op_{2}$\nf{.Append()} started. From Lemma \ref{lem::append} $op_{1}$ is in \nf{root.blocks} before $op_{2}$ starts to propagate. By definition of $L$, $op_{1}$ is linearized before $op_{2}$.
\end{proof}
Once some operations are aggregated in one block, they will get propagated up to the root together and they can be linearized in any order among themselves. The way we have chosen to order the operations in a block makes \nf{GetEnqueue} and \nf{FindResponse} efficient.

\begin{definition}\label{defNullDeq}
If a \nf{Dequeue} operation returns null it is called a \it{null dequeue}, otherwise it is called \it{non-null dequeue}.
\end{definition}

\begin{definition}
   Assume the operations in \nf{root.blocks} are applied sequentially on an empty queue in the order of $L$.  ${Resp(d)=e}$\nf{.element} if the element of \nf{Enqueue} $e$ is the response to \nf{Dequeue} $d$. Otherwise if ${d}$ is a null dequeue then ${Resp(d)=\nf{null}}$. 
\end{definition}
